Rheology of Cubic Blue Phases

### 2012-09-11 papers/bprheo/bprheo.tex
Update bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1782 40876441-3496-0410-a613-a2fca35b2ae2

In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
phases BPI and BPII. The thermodynamic states are characterised by an effective temperature 
$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy terms.
We opted for states sufficiently far away from the chiral-nematic-isotropic 
where both quiescent blue phases are at equilibrium, namely
$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ (BPII), respectively.
Simulations at metastable states at higher and lower temperatures and different chiralities showed 
no differences in their generic flow behaviour.
However, states closer to the phase boundary featuring higher free energy densities and 
smaller average order parameters show a tendency to break up more easily under external driving forces.

As is common in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our simulations with 
analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, 
and equilibrated these (for 5000 LB timesteps) prior to starting the shear. 
During the equilibration run the optimal redshift $r$ was recalculated at every timestep.
After equilibration the redshift parameter was kept constant throughout the rest of the simulation.
We chose a unit cell size of 16 lattice sites for BPII and 32 sites in case of BPI. 
Runs with higher resolution confirmed that this was an appropriate minimal resolution to cover 
the complex conformational changes and kinematic details of the blue phase networks.

Simple shear flow is imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) half of the planes moving in positive (negative) x-direction.
We applied a correction for spurious momentum currents that emerge due to interpolations accross 
the Lees-Edwards planes \cite{Henrich:2012a}.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $6.3\times\e{-4}$ in LB units.
The shear flow was imposed along the x-axis, whereas velocity gradient and vorticity are oriented 
along the y- and z-axis.
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
is equal to about $10^8$~Pa.
More details about the conversion from LBU to SI units can be found in
\cite{Henrich:2011a,Henrich:2010b}.

@@

We start the discussion with BPII, which shows comparably simpler flow behaviour than BPI.
BPII features simple cubic symmetry and consists of $\pi$-disclination lines that 
intersect and form a characteristic network of nodes.

A quantitative analysis of the flow behaviour of BPII is shown in Fig. \ref{bp2-rheo}.
The apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ over time normalised to
the isotropic background viscosity $\eta_{app}$ is given in the left image, whereas
the picture on the right shows the same data versus accummulated total strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is due to the periodic break-up and reconnection of the network under shear flow.

The inset shows the time average of the deviatoric part of the stress tensor $\bar{Pi}_{xy}$ as a function of shear rate.
The error bars indicate the maximum and minimum stresses that occur during one cycle.
The data for low and intermediate $\gd$ obeys a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.34(99), b=0.95(38)$ to
a very good approximation, exibiting a remarkably small degree of shear-thinning.
This probably reflects the very regular and rather undisturbed affine transformation of the disclination network.

\begin{figure}[ht!]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
\label{bp2-rheo}
\end{figure}

\subsubsection{Low and intermediate shear rates}

Fig \ref{bp2-med} shows the disclination network 
undergoing an affine transformation under shear. The disclination lines break up and 
reconnect further downstream forming a periodically recurring pattern. 
The recurrence time in flow direction is $\tau_F=p/(2\, v_{LE})=p/(2\,\gd\, \Delta y_{LE})= 1/\gd$.
Apart from the affine transformation the general appearance is very close to that 
of the quiescent equilibrium phase.

\begin{figure}[ht]
\includegraphics[width=0.4\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.4\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.4\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
of shear-induced transformations in the steady state at 
$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
The boundary conditions impose flow in such a way that network moves 
to the right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure}

Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
the gradient direction in the direction of vorticity.
Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
making the network travel along the z-direction.
The recurrence period in vorticity direction is exactly six times larger than that of reconnecting in flow direction.

Fig. \ref{bp2-velo} shows the disclination network at a characteristic time step between reconnecting and another breakup.
Superimposed are vectors which represent the velocity projected onto a plain perpendicular to the flow direction.
This projection makes it possible to get rid of the dominating velocity component $v_x$ and permits visualisation the 
two much smaller components $v_y$ and $v_z$.
The magnitude of the minor components is typically in the range of a few percent of the primary flow component.
Characteristic bands oriented along the vorticity direction form, which are normally hidden behind the primary flow in x-direction.
When the helicity of the underlying cholesteric phase is changed from a left- to a righthanded helix the sign of $v_z$ 
and the sense of motion also changes.
Further quantitative evidence for a direct link between flow direction and helicity can be found by time-averaging over
individual cycles.
Tab. \ref{tab1} gives minima, maxima, averages and standard deviations of the velocity components.
All values for the two runs with inverted helicity are identical apart from a change of sign in the z-components.
There is only a slight imbalance between the maximum and minimum velocity in z-direction.
Hence the movement of the network in vorticity direction is predominantly permeative.
Note that the imbalance is exaclty retained if the helicity is inverted.

\begin{figure*}[ht]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with positive helicity travels rightwards, whereas the one one the right
moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
to reconnect in flow direction.}
\label{bp2-velo}
\end{figure*}

\subsubsection{High shear rates}

If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation under shear that we
descibed above. Fig. \ref{bp2-high} shows a cut through the director field of BPII at high shear rate.
The entire blue phase network has breaken up into a simple cholesteric phase. The preferred mode of flow is 
a travelling helical wave with the helical axis oriented along the vorticity direction and translational 
invariance in flow and gradient direction. This mode of flow is the preferred one for low Ericksen numbers
before a helix uncoiling transition to a nematic flow aligned state takes place at even higher
shear rates \cite{Rey:1996a, Rey:1996b}.

In our scenario the break-up occurs for shear rates between $\gd=3.12\e{-4}$ and $6.25\e{-4}$.
However, it is reasonable to assume that the critical shear rate depends on thermodynamic and kinetic parameters
such as the effective temperature $\tau$, the chirality $\kappa$ and the rotational diffusion constant $\Gamma$.
Due to a lack of computational resources we have not investigated these aspects any further.

\begin{figure*}[h]
\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. The configuration is a helical wave that travels along the direction of vorticity. It is translationally invariant in gradient and flow direction.}
\label{bp2-high}
\end{figure*}

@@

In BPI the disclination lines are arranged in a body-centered-cubic array.
Contrary to BPII, they are well separated and do not intersect. 
There is strong evidence that these topological differences result in 
a very different rheological response of BPI. In fact BPI exhibits some 
peculiarities that are absent in BPII.

Just as before in a kind of overview we present rheological key quantities first and 
will address the detail aspects further below. 
Fig. \ref{bp1-rheo} shows the apparent viscosity over time (left graph) and 
strain (right graph) for intermediate and high shear rates. In case of BPI the oscillatory patterns
are no longer symmetric as it takes more time to build up stress and reach a maximum than to 
relax. Shear-thinning is much more pronounced than for BPII. At the lowest shear rate $\gd=9.76\e{-6}$
the apparent viscosity varies by almost a factor 8 during one cycle.
The decrease in apparent viscosity is monontonous for decreasing shear rate, but does not 
follow a simple power law.

The inset of shows the flow curve, i.e. shear stress over shear rate.
The data points are averages over one cycle, whereas the error bars indicate the maximum and
minmum values during one cycle. Depending on the shear rate three different regimes of flow emerge:
At high shear rates the network breaks up similar to BPII. The director field in this state
is effectively two-dimensional and translationally invariant along the neutral direction.
At intermediate shear rates the BPI network flows with periodically recurring conformations.
These complex flow-induced conformations are accompanied by large fluctuations that increase 
in intensity for decreasing shear rate. At low shear rates the minimum shear stress contribution
of the liquid crystal during one cycle is actually negative. 
Initially, the network undergoes an affine transformation, but it becomes unstable at larger stages. 
Finally it dissolves into an amorphous configuration that has separated disclination lines. This 
end state is the same for all shear rates below a critical threshold, which was below $\gd=9.76\e{-5}$.

\begin{figure}[ht]
\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. Three different regimes can be identified: amorphous BP network with yield stress (triangles), flow with recurring oscillatory patterns (open circles), breakup (solid circle). The error bars give the minimum and maximum shear stress during one cycle.}
\label{bp1-rheo}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI at different shear rates: The main picture shows the frequency shift of the Fourier components in agreement with the imposed shear rate. The inset depicts the same analysis at higher shear rate, where the frequency shift is no longer observed. Instead the spectral contribution of higher modes grows until finally the breakup takes place.}
\label{bp1-spectrum}
\end{figure}

\subsubsection{Low shear rates}

The rheological response of BPI at low shear rate is strikingly 
different from that of BPII.
We want to explain these differences on the basis of a typical 
appearance of BPI at low flow rate, which is depicted in Fig. \ref{bp1-low}.
Three different stages can be distinguished. 
With the onset of shearing the disclination lines 
in BPI get more and more squeezed together which is incommensurable with 
the defect topology of this blue phase. 
Consequently, the network adopts a flow-induced conformation which consists 
of intertwined helices that stretch during the affine transformation.
This helical conformation that emerges already at strains $\gamma\simeq1$ 
is shown in Fig. \ref{bp1-low} (top row).
At a particular point in the cycle a non-helical configuration is
observed (bottom left) before the network quickly returns
to its helical state.
There is no perceivable movement of the network in vorticity 
direction at any stage of the cycle.

However, this mode of flow proves unstable as after a few cycles 
distortions appear which lead to further destabilisation.
Finally, the network

\begin{figure*}[ht]
\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
\caption{Snapshots of BPI disclination network at low shear rate ($\dot{\gamma}=4.88\e{-6}$): Depicted is the transition from flow-induced, intertwined helices that undergo a recurring affine transformation (top left and centre at time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) to a distorted state (bottom centre at time step $t=1.7\e{6}$) and finally to an amorphous network (bottom right at $t=2.2\e{6}$).}
\label{bp1-low}
\end{figure*}

\begin{figure}[ht]
\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
\caption{}
\label{bp1-fe-yield}
\end{figure}

\subsubsection{Intermediate shear rates}

Fig. \ref{bp1-med} shows a typical sequence of shear-induced conformations.
The recurrence time in flow direction is $\tau_F=L_x/(2\, v_{LE})=L_x/(2\,\gd\, \Delta y_{LE})= 4/\gd$.

There is again an offset in the position of the network at the begin and the end of each cycle. 
But neither the magnitude nor the direction of the offset seem to obey a clear rule.
They are seemingly chaotic and different for every shear rate.

>>>>

\begin{table*}[htpb]
\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
\hline
$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
\hline
BPI \\
\hline
0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
\hline
BPII \\
\hline
0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
\hline
\end{tabular}
\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding any transient dynamics.}
\label{tab1}
\end{table*}

In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
phases BPI and BPII. 
The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
is equal to about $10^8$~Pa. Further details about the conversion 
from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.

The thermodynamic states are characterised by an effective temperature 
$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy.
For the vast majority of simulations we opted for equilibrium states that are sufficiently 
far away from the cholesteric-isotropic transition, namely
$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ (BPII), respectively.
We performed also a number of simulations at metastable states at higher and lower temperatures 
and different at different chiralities, but didn't find any significant differences in the 
generic flow behaviour that is described below.
However, for thermodynamic states that are closer to the phase boundary the disclination 
network broke up into non-BP states at lower shear rates. This is not surprising as
these states have on average higher free energy densities and smaller order parameters
than those specified above and can therefore not withstand the same external forces.

As usual in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our runs with 
analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit 
and equilibrated these configuration for 5000 LB timesteps before we started the shear flow. 
During the equilibration sequence the optimal redshift $r$ was calculated and applied at every timestep.
After equilibration the redshift was kept constant throughout the rest of the simulation with shear flow.
We chose a pitch length of 32 LBU for BPII and 64 LBU in case of BPI, which is the equivalent 
of 4 unit cells along each coordinate direction and a total of 64 unit cells in our simulation box.
Runs with higher resolution confirmed that the resolution was adequate to track  
all kinematic details of the blue phase networks in shear flow.

Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) half of the planes moving in positive (negative) x-direction and the 
velocity gradient along the y-direction.
We applied a correction for spurious momentum currents that emerge due to required interpolations across 
the Lees-Edwards planes \cite{Henrich:2012a}.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
For clarity we classify various subregimes, two in case of BPII and three in case of BPI. 
They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (top) and strain (bottom). For clarity the individual curves have been shifted along the time and strain axis. The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: regular periodic breakup and reconnecting of the discination lines (BPII-1, open circles) and break up of the network into a travelling helical wave (BPII-2, solid circle).}
\label{bp2-rheo}
\end{figure}

We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
than that of BPI. BPII has simple cubic symmetry and the $\pi$-disclination lines intersect and form a characteristic network of nodes.

As it might be easier to enter a discussion with the full picture in mind, we provide first a 
quantitative analysis of the flow behaviour at all applied shear rates.
Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
deviatoric stress normalised to the isotropic background viscosity: 
$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
A numerical value of $\eta_{app}=1$ corresponds to total shear thinning without 
any additional contribution from the liquid crystal.
The top picutre shows the date over time, whereas the graph at the bottom gives 
the same data versus accummulated strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is entirely because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.

The inset shows a kind of flow curve, defined as time averages of 
the deviatoric stress tensor $\bar{Pi}_{xy}$ as a function of shear rate $\gd$.
The error bars indicate the maximum and minimum stresses that occur during one cycle.
For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
$a=0.35, b=0.95$ describes the data to a very good approximation. 
Hence, the degree of shear-thinning is remarkably small, but may depend
on the exact direction in which the strain is applied, an aspect that we     
leave for future work.

For shear rates $\gd\ge6.25\e{-4}$ the network of disclination lines
breaks up and oscillations in the stress signal are absent.
In this regime of shear rates, to which we refer as BPII-2, the 
blue phase dissolves into a simple cholesteric liquid 
crystal. The mode of flow is a travelling helical wave with the helical axis 
oriented along the vorticity direction.

In the next paragraph we will investigate this in more detail by looking at the 
kinetics of the disclination network.

\subsubsection{Regime BPII-1: low and intermediate shear rates}

Fig. \ref{bp2-med} shows the disclination network in shear flow as
it undergoes an affine transformation. The disclination lines break up and 
reconnect further downstream forming a periodically recurring pattern with 
recurrence time $\tau_F=1/\gd$ in flow direction.

\begin{figure}[htpb]
\includegraphics[width=0.4\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.4\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.4\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: 
The pictures show a typical sequence of snapshots in the steady state 
at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
gradient is oriented along the vertical direction, whereas the 
horizontal direction is the flow direction. Lees-Edwards boundary 
conditions have been imposed in such a way that network moves to the 
right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure}

\begin{figure*}[htpb]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with positive helicity travels rightwards, whereas the one one the right
moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network to reconnect in flow direction.}
\label{bp2-velo}
\end{figure*}

The general appearance of the flowing network is apart from the affine transformation 
very close to that of the quiescent blue phase at equilibrium. It is the same for 
all shear rates in regime BPII-1 and a first explanation why shear-thinning is so weak.

Interestingly, while being displaced with the flow the entire network moves perpendicular 
to the flow and velocity gradient in the vorticity direction.
A similar behaviour has been recently observed for blue phases in shear flow and 
confined geometries ~\cite{Henrich:2012b}.
But contrary to the stick-slip motion that has been reported there the movement is 
steady. It occurs in such a manner that the positions of breakup and reconnection 
of the network, visible in Fig. \ref{bp2-med}, are slightly offset and make the network 
travel along the z-direction.
The recurrence time in vorticity direction $\tau_V=6\tau_F$, 
i.e. it takes a displacement of six unit cells in flow directions 
until the network has moved one unit cell in voriticity direction.

Fig. \ref{bp2-velo} shows the disclination network sometime between reconnecting 
and the next breakup with superimposed velocity vectors.
Shown are what we may call secondary velocity components, obtained by projecting the
velocity onto a plain perpendicular to the flow direction.
This allows to get rid of the dominating velocity component $v_x$ and to
visualise the two much smaller components $v_y$ and $v_z$.
The magnitude of the secondary components is typically in the range of a 
few percent of the primary flow component.

Characteristic bands are visible, which are oriented along the vorticity direction and
normally hidden behind the primary flow component.
When the helicity of the underlying cholesteric phase is changed from left- to a 
righthanded the pattern becomes a mirror image of the former one. The sign of 
$v_z$ change and also the sense of motion is inverted.
Further quantitative evidence for a direct link between the sense of motion 
and the helicity can be gained by time-averaging over individual cycles.
Tab. \ref{tab1} gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the z-components.
There is only a slight imbalance between the maximum and minimum velocities 
in z-direction, suggesting that the movement of the network in vorticity 
direction is permeative. Note that this imbalance is retained under inversion of
helicity.

\subsubsection{Regime BPII-2: high shear rates}

If the flow rate exceeds a certain critical value BPII does not undergo the 
affine transformation descibed above and adopts another mode of flow. 
Fig. \ref{bp2-high} shows a sequence of cuts through the director 
field at high shear rate. The entire blue phase network has broken up 
into a simple cholesteric helix and flows as a travelling 
helical wave with the helical axis oriented along the vorticity direction.
The state is translationally invariant in flow and gradient direction.
and is the preferred mode of flow of a cholesteric 
liquid crystal at low Ericksen numbers below the uncoiling transition 
to a nematic, flow-aligned state at even higher shear rates ~\cite{Rey:1996a, Rey:1996b}.

In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$.
At the moment we do not have a hard criterium for the critical
shear rate at hand. However, it seems reasonable that the critical shear rate 
depends on thermodynamic and kinetic parameters such as the effective 
temperature $\tau$, the chirality $\kappa$ and the rotational diffusion 
constant $\Gamma$.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. 
The configuration is a helical wave that travels along the direction of vorticity.
It is translationally invariant in gradient and flow direction.}
\label{bp2-high}
\end{figure}

@@

BPI has body-centered symmetry and contrary to BPII the disclination lines
 are well separated and do not intersect.
We believe that this topological difference is responsible for most of
the differences between BPI and BPII regarding their rheological response.

Again, we present key aspects in a kind of overview before we 
address some features in more detail. Fig. \ref{bp1-rheo} shows the 
apparent viscosity $\eta_{app}$ over time and strain for the same shear rates
as in Fig. \ref{bp2-rheo}.
For shear rates in regime BPI-3, i.e. $\gd\ge6.25\e{-4}$ a breakup of 
the disclination network occurs similar to and at the same critical shear rate as 
the one observed in BPII-2 but with an order structure that is a quasi 
two-dimensional.

In regime BPI-2 at lower shear rates periodic oscillations emerge in the 
deviatoric stress that are not sinusoidal, but show an increasingly complex
pattern up to the breakup point.
For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
they become  unregular and finally disappear after a few
cycles. Shear-thinning is slightly stronger than in BPII. 
The average shear stress during one cycle can be fitted reasonably 
well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (top) 
and strain (bottom). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. 
Depending on the configuration in steady shear flow three different regimes 
can be identified: amorphous BP network with yield stress (BPI-1, triangles), 
steady flow with recurring patterns (BPI-2, open circles) and 
breakup of the disclination network (BPI-3, solid circle). 
The error bars represent the minimum and maximum shear stress 
during one cycle. For clarity the individual curves have been shifted along
the time and strain axis.}
\label{bp1-rheo}
\end{figure}

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI at different shear rates: The main picture shows the frequency shift of the Fourier components in agreement with the imposed shear rate. The inset depicts the same analysis at higher shear rate, where the frequency shift is no longer observed. Instead the spectral contribution of higher modes grows until finally the breakup takes place. The frequency is given in units of $10^2\cdot 2\pi/T$ with $T$ as period of one cycle in LB time steps.}
\label{bp1-spectrum}
\end{figure}

\subsubsection{Regime BPI-1: low shear rates}

The rheological response of BPI at low shear rate which exhibits a
dynamic yield stress is strikingly different from that of BPII.
An explanation for this strange behaviour can be found by 
looking more closely at the average deviatoric stress and 
free energy density, as shown in Fig. \ref{bp1-fe-yield}.
When the quiescent and equilibrated BPI network begins to flow
the shear stress increases steeply and goes through a global maximum.
Shortly after it goes negative (the total shear stress is still
positive as the backgound viscosity has been subracted) and positive
again, forming a complete cycle. 
This cycle is repeated for a couple of times before 
the amorphous configuration emerges and any regular pattern
disappears.
The negative branch in the deviatoric stress is connected
to a local maximum followed by a local minimum in the average free energy.
Hence, the region of negative stress can be 
regarded as a phase during which elastic forces excert a 
'pull' onto the network while it tries to reach a more 
favourable configuration. However, the low flow rate
causes this unstable intermediate state to be exposed to these
pulling elastic forces for a relatively long time, long enough to 
create small distortions which finally break the symmetry of the
flow-induced configuration.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
\caption{Average deviatoric stress and free energy density of BPI at low shear rate: The negative branch in the stress is related to a local maximum and a following local minimum in free energy density. This creates unstable conditions that lead to a dissolving regular network and an amorphous state which features a yield stress.}
\label{bp1-fe-yield}
\end{figure}

We want to explain these differences on the basis of a typical 
appearance of BPI at low flow rate, which is depicted in Fig. \ref{bp1-low}.
Three different stages can be distinguished. 
With the onset of shearing the disclination lines 
in BPI get more and more squeezed together which is incommensurable with 
the defect topology of this blue phase. 
Consequently, the network adopts a flow-induced conformation which consists 
of intertwined helices that stretch during the affine transformation.
This helical conformation that emerges already at strains $\gamma\simeq1$ 
is shown in Fig. \ref{bp1-low} (top row).
At a particular point in the cycle a non-helical configuration is
observed (bottom left) before the network quickly returns
to its helical state.
There is no perceivable movement of the network in vorticity 
direction at any stage of the cycle.
